---
title: "Predicting Pasture Growth with SSURGO data"
author: "Elissa Chasen"
date: "4/2/2020"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = "/Volumes/GoogleDrive/My Drive/grassland2.0/grazescape/Soil Data/SSURGO data/")
```

## Predicting grass growth based on soil characteristics

* Depending on the year, soil properties and topography can explain 60% or more of crop yield variability (Jiang & Thelan, 2004)
* SSURGO (Soil Survey Geographic Database) provides representative yields for several grass-legume (pasture) crops associated with soil components.
* Each soil component:
  + can be geographically mapped by connecting the component to the mapunit
  + is associated with many representative soil charactistics (slope, % silt, cec, etc.)


## Predicting grass growth based on soil characteristics

* We first download and clean the soil for our geographic region of interest.
* See: https://rpubs.com/emchasen/SSURGOcleaning

## Create working environment
Built with R version `r getRversion()`
```{r}
#load libraries
library(dplyr)
library(ggplot2)
library(tidyr)
library(GGally)
library(caret)
library(MASS)
library(leaps)
library(gt)
library(car)
```

## Representative component crop yields

```{r}
crawford_crop <- 
  read.csv("EXTRACTIONS/Crawford/SSURGO/Crawford_SSURGO_cocropyld.csv")
vernon_crop <- 
  read.csv("EXTRACTIONS/Vernon/SSURGO/Vernon_SSURGO_cocropyld.csv")
# combine each county to make full data set
crop <- rbind(crawford_crop, vernon_crop)
colnames(crop)
crop <- crop %>% dplyr::select(c(cropname, nonirryield.r, cokey)) %>%
  rename(yield = nonirryield.r) %>%
  mutate_if(is.character, as.factor)
```

## Examine data

```{r}
levels(crop$cropname)
```

## Examine data

We are only interested in grass-legume combinations.

```{r}
grass <- crop %>%
  filter(cropname == "Bluegrass-white clover"|
           cropname == "Bromegrass-alfalfa hay"|
           cropname == "Grass-clover"|
           cropname == "Grass-legume pasture"|
           cropname == "Orchardgrass-alsike"|
           cropname == "Orchardgrass-red clover"|
           cropname == "Timothy-alsike") %>%
  droplevels()
```

## Examine data
Now we have 7 different grass-legume combinations. Summary data of each. 
```{r, echo=FALSE}
sum_grass <- grass %>%
  group_by(cropname) %>%
  summarize(min = min(yield, na.rm = TRUE),
            mean = mean(yield, na.rm = TRUE),
            max = max(yield, na.rm = TRUE),
            count = n()) %>%
  arrange(desc(count))  %>%
  gt() %>%
  tab_header(
    title = "Representative yields (tons/acre)",
    subtitle = "Crawford and Vernon County") %>%
  cols_label(
    cropname = md("**Grass mixture**"),
    min = md("**Min**"),
    mean = md("**Mean**"),
    max = md("**Max**"),
    count = md("**No. Obs**")
  )
sum_grass
```

## Let's remove the crops without many data points. 
They are mostly anomolous, and the grass-legume mixture doesn't add any value.
```{r, echo = FALSE}
grass <- grass %>%
  filter(cropname == "Bluegrass-white clover"|
           cropname == "Orchardgrass-alsike"|
           cropname == "Orchardgrass-red clover"|
           cropname == "Timothy-alsike") %>%
  droplevels()

sum_grass <- grass %>%
  group_by(cropname) %>%
  summarize(min = min(yield, na.rm = TRUE),
            mean = round(mean(yield, na.rm = TRUE),2),
            max = max(yield, na.rm = TRUE),
            count  = n()) %>%
  arrange(desc(count)) %>%
  gt() %>%
  tab_header(
    title = "Representative yields (tons/acre)",
    subtitle = "Crawford and Vernon County") %>%
  cols_label(
    cropname = md("**Grass mixture**"),
    min = md("**Min**"),
    mean = md("**Mean**"),
    max = md("**Max**"),
    count = md("**No. Obs**")
  )
sum_grass
```           

## Yield distribution
```{r, echo=FALSE}
ggplot(grass, aes(x=yield)) +
  geom_histogram(alpha=0.8, position = 'identity') +
  xlab("Yield (tons/acre)")
```

## Distribution of yields by crop

```{r, echo=FALSE}
grassYield <- ggplot(grass, aes(x=yield, fill=cropname)) +
  geom_histogram(alpha=0.6, position = 'identity') +
  facet_wrap(~cropname)+
  xlab("Yield (tons/acre)")+
  theme(legend.position = "none")
grassYield
```

## Add grass yields to cleaned soil data

```{r}
#upload soils file
soil <- read.table("Final Clean Soil/fullSoilCrawfordVernon.txt", 
                   header = TRUE, sep = ",")
#summary(soil)
#remove high OM (<55) mucky soils 
soil <- soil %>%
  filter(om < 55)
#join crop yield and soil data by cokey
grassSoil <- left_join(grass, soil, by = "cokey")
# some cokey's from the yield data do not have matching cokeys from the soil data
# remove yield observations with no corresponding soil data (was from mucky soils)
grassSoil <- grassSoil %>%
  drop_na()
```

## Variable selection

Step 1. Create training and test sets of data

```{r, cache=TRUE}
set.seed(0731)
inTrain <- createDataPartition(grassSoil$yield, p = 0.75, list = FALSE)
trainGM <- grassSoil[inTrain,]
testGM <- grassSoil[-inTrain,]
```

## Variable selection

Step 2. Null model hypothesis

* This gives us a starting place to compare other models

```{r}
best.guess <- round(mean(trainGM$yield),2) 
# Evaluate RMSE
RMSE.baseline <- round(sqrt(mean((best.guess-testGM$yield)^2)),2)
```
The average yield of all of the grass varieties across all of the variables is `r best.guess` and the difference between that average and the error based on untrained data is `r RMSE.baseline`.

## Variable selection: Slope

Slope directly affects crop growth through changing and redirecting water availability. It indirectly effects crop growth by changing other soil properties (Jiang & Thelan, 2004).

```{r,echo=FALSE}
#quantile(grassSoil$slope.r, probs = seq(0,1,0.2))# set up cut-off values 
breaks <- c(1,2,8,16,25,70)
# specify interval/bin labels
tags <- c("1-2","2-8", "8-16", "16-25", "25-70")
# bucketing values into bins
group_tags <- cut(grassSoil$slope.r, 
                  breaks=breaks, 
                  include.lowest=TRUE, 
                  right=FALSE, 
                  labels=tags)
grassSoil$slopeCat <- group_tags

sum_slope <- grassSoil %>%
  group_by(slopeCat) %>%
  summarize(mnYld = mean(yield, na.rm = TRUE),
            stdev = sd(yield, na.rm = TRUE),
            count = n(),
            se = stdev/sqrt(count))

ggplot(sum_slope, aes(x = slopeCat, y = mnYld, label = count))+
  geom_point(size = 3)+
  geom_text(vjust = -2, (aes(label = paste("n =", count))))+
  ylim(2,4)+
  geom_errorbar(aes(ymin = mnYld - se, ymax = mnYld + se, width = 0.2))+
  ylab("Mean Yield (tons/acre) +/- se")+
  xlab("Slope %")+
  ggtitle("Grass yield by slope percent (binned into 5 quantiles)")
```

## Variable selection: organic matter

```{r, echo = FALSE}
quantile(grassSoil$om, probs = seq(0,1,0.25))# set up cut-off values
breaks <- c(0,2,4,6, 8)
#specify interval/bin labels
tags <- c("0-2","2-4", "4-6", "6-8")
#bucketing values into bins
group_tags <- cut(grassSoil$om,
                breaks=breaks,
               include.lowest=TRUE,
              right=FALSE,
             labels=tags)
#summary(group_tags)
grassSoil$omCat <- group_tags
sum_om <- grassSoil %>%
group_by(omCat) %>%
summarize(mnYld = mean(yield, na.rm = TRUE),
        stdev = sd(yield, na.rm = TRUE),
       count = n(),
      se = stdev/sqrt(count))
#sum_om
ggplot(sum_om, aes(x = omCat, y = mnYld))+
geom_point(size = 3)+
 geom_errorbar(aes(ymin = mnYld - se, ymax = mnYld + se, width = 0.5))+
ylab("Mean Yield (tons/acre) +/- se")+
xlab("Organic Matter")+
ggtitle("Grass yield by organic matter")

ggplot(grassSoil, aes(x = om, y = yield))+
  geom_point() +
  xlab("Organic Matter")+
  ylab("Yield (tons/acre)")+
  ggtitle("Relationship between yield and OM")
```

## Organic matter

```{r, echo = FALSE}
om.mod <- lm(yield~om, data = trainGM)
pred.om <- predict(om.mod, newdata = testGM)
RMSE.om <- round(sqrt(mean((pred.om-testGM$yield)^2)),2)
rsquared.om <- round((summary(om.mod)$r.squared),2)
```
Organic matter predicts approx `r rsquared.om*100` percent of the variability in yield. 

```{r, echo = FALSE}
plot(pred.om~testGM$yield, ylab = "Predicted yield from OM model", xlab = "Actual yield from test dataset") 
abline(a = 0, b = 1)
```

## Variable selection: Cation Exchange Capacity

```{r, echo = FALSE}
#quantile(grassSoil$cec, probs = seq(0,1,0.2))# set up cut-off values 
breaks <- c(1.65,5.73,9.37,13.30,14.83,20.01)
# specify interval/bin labels
tags <- c("1.65-5.73","5.73-9.37", "9.37-13.30", "13.30-14.83", "14.83-20.01")
# bucketing values into bins
group_tags <- cut(grassSoil$cec, 
                  breaks=breaks, 
                  include.lowest=TRUE, 
                  right=FALSE, 
                  labels=tags)
#summary(group_tags)
grassSoil$cecCat <- group_tags
sum_cec <- grassSoil %>%
  group_by(cecCat) %>%
  summarize(mnYld = mean(yield, na.rm = TRUE),
            stdev = sd(yield, na.rm = TRUE),
            count = n(),
            se = stdev/sqrt(count))
#sum_cec
ggplot(sum_cec, aes(x = cecCat, y = mnYld, label = count))+
  geom_point(size = 3)+
  geom_text(vjust = -2, (aes(label = paste("n =", count))))+
  geom_errorbar(aes(ymin = mnYld - se, ymax = mnYld + se, width = 0.2))+
  ylim(1.5,4.2)+
  ylab("Mean Yield (tons/acre) +/- se")+
  xlab("Cation exchange capacity")+
  ggtitle("Grass yield by CEC (binned into 5 quantiles)")

#ggplot(grassSoil, aes(x = cec, y = yield))+
 # geom_point()
```

## Cation exchange capacity

```{r, echo = FALSE}
cec.mod <- lm(yield~cec, data = trainGM)
pred.cec <- predict(cec.mod, newdata = testGM)
RMSE.cec <- round(sqrt(mean((pred.cec-testGM$yield)^2)),2)
rsquared.cec <- round((summary(cec.mod)$r.squared),2)
```
Cation exchange capacity predicts approx `r rsquared.cec*100` percent of the variability in yield. 

```{r, echo = FALSE}
plot(pred.cec~testGM$yield, ylab = "Predicted yield from CEC model", xlab = "Actual yield from test dataset") 
abline(a = 0, b = 1)
```

## Variable selection: Available Water Capacity

```{r, echo = FALSE}
#quantile(grassSoil$awc, probs = seq(0,1,0.25), na.rm = TRUE)# set up cut-off values 
breaks <- c(0.07,0.14,0.18,0.22,0.26)
# specify interval/bin labels
tags <- c("0.07-0.14","0.14-0.18", "0.18-0.22", "0.22-0.26")
# bucketing values into bins
group_tags <- cut(grassSoil$awc, 
                  breaks=breaks, 
                  include.lowest=TRUE, 
                  right=FALSE, 
                  labels=tags)
#summary(group_tags)
grassSoil$awcCat <- group_tags
sum_awc <- grassSoil %>%
  group_by(awcCat) %>%
  summarize(mnYld = mean(yield, na.rm = TRUE),
            stdev = sd(yield, na.rm = TRUE),
            count = n(),
            se = stdev/sqrt(count))
#sum_cec
ggplot(sum_awc, aes(x = awcCat, y = mnYld, label = count))+
  geom_point(size = 3)+
  geom_errorbar(aes(ymin = mnYld - se, ymax = mnYld + se, width = 0.2))+
  geom_text(vjust = -2, (aes(label = paste("n =", count))))+
  ylim(1.5, 4.1)+
  ylab("Mean Yield (tons/acre) +/- se")+
  xlab("Available water capacity")+
  ggtitle("Grass yield by AWC (binned into 4 quantiles)")
```

## Available Water Capacity

```{r, echo=FALSE}
awc.mod <- lm(yield~awc, data = trainGM)
pred.awc <- predict(awc.mod, newdata = testGM)
RMSE.awc <- round(sqrt(mean((pred.awc-testGM$yield)^2)),2)
rsquared.awc <- round((summary(awc.mod)$r.squared),2)
```
Available water capacity predicts approx `r rsquared.awc*100` percent of the variability in yield. 

```{r, echo = FALSE}
plot(pred.awc~testGM$yield, ylab = "Predicted yield from AWC model", xlab = "Actual yield from test dataset") 
abline(a = 0, b = 1)
```


## Variable Selection: Saturated hydraulic conductivity (ksat)

```{r, echo = FALSE}
#hist(grassSoil$ksat, breaks = 30)
#summary(grassSoil$ksat)
#quantile(grassSoil$ksat, probs = seq(0,1,0.25))# set up cut-off values 
breaks <- c(6.54,9.17,12.27, 24,181.14)
# specify interval/bin labels
tags <- c("6.54-9.17","9.17-12.27", "12.27-24", "24-181.14")
# bucketing values into bins
group_tags <- cut(grassSoil$ksat, 
                  breaks=breaks, 
                  include.lowest=TRUE, 
                  right=FALSE, 
                  labels=tags)
#summary(group_tags)
grassSoil$ksatCat <- group_tags
sum_ksat <- grassSoil %>%
  group_by(ksatCat) %>%
  summarize(mnYld = mean(yield, na.rm = TRUE),
            stdev = sd(yield, na.rm = TRUE),
            count = n(),
            se = stdev/sqrt(count))
#sum_ksat
ggplot(sum_ksat, aes(x = ksatCat, y = mnYld, label = count))+
  geom_point(size = 3)+
  geom_text(vjust = -2, (aes(label = paste("n =", count))))+
  geom_errorbar(aes(ymin = mnYld - se, ymax = mnYld + se, width = 0.2))+
  ylim(1.5,4.1)+
  ylab("Mean Yield (tons/acre) +/- se")+
  xlab("Saturated hydraulic conductivity")+
  ggtitle("Grass yield by ksat (binned)")
```

## Raw data: ksat 

```{r, echo = FALSE}
ggplot(grassSoil, aes(x = ksat, y = yield))+
  geom_point()+
  xlab("Saturated hydraulic conductivity")+
  ylab("Yield, tons/acre")
```

## ksat

```{r, echo = FALSE}
ksat.mod <- lm(yield~ksat, data = trainGM)
pred.ksat <- predict(ksat.mod, newdata = testGM)
RMSE.ksat <- round(sqrt(mean((pred.ksat-testGM$yield)^2)),2)
rsquared.ksat <-  round((summary(ksat.mod)$r.squared),2)
```
Saturated hydraulic conductivity predicts approx `r rsquared.ksat*100` percent of the variability in yield. 

```{r, echo = FALSE}
plot(pred.ksat~testGM$yield, ylab = "Predicted yield from Ksat model", xlab = "Actual yield from test dataset") 
abline(a = 0, b = 1)
```

## Variable selection: pH

```{r, echo = FALSE}
#hist(grassSoil$ph, breaks = 30)
#summary(grassSoil$ph)
#quantile(grassSoil$ph, probs = seq(0,1,0.2), na.rm = TRUE)# set up cut-off values 
breaks <- c(5.37,5.88,6.2,6.3,7.7)
# specify interval/bin labels
tags <- c("5.37-5.88","5.88-6.2", "6.2-6.3", "6.3-7.7")
# bucketing values into bins
group_tags <- cut(grassSoil$ph, 
                  breaks=breaks, 
                  include.lowest=TRUE, 
                  right=FALSE, 
                  labels=tags)
#summary(group_tags)
grassSoil$phCat <- group_tags
sum_ph <- grassSoil %>%
  group_by(phCat) %>%
  summarize(mnYld = mean(yield, na.rm = TRUE),
            stdev = sd(yield, na.rm = TRUE),
            count = n(),
            se = stdev/sqrt(count))
#sum_ph
ggplot(sum_ph, aes(x = phCat, y = mnYld, label = count))+
  geom_point(size = 3)+
  geom_text(vjust = -2, (aes(label = paste("n =", count))))+
  geom_errorbar(aes(ymin = mnYld - se, ymax = mnYld + se, width = 0.2))+
  ylab("Mean Yield (tons/acre) +/- se")+
  xlab("pH")+
  ylim(2.1,4.2)+
  ggtitle("Grass yield by pH (binned)")

```

## pH

```{r, echo = FALSE}
ph.mod <- lm(yield~ph, data = trainGM)
pred.ph <- predict(ph.mod, newdata = testGM)
RMSE.ph <- round(sqrt(mean((pred.ph-testGM$yield)^2)),2)
rsquared.ph <- round((summary(ph.mod)$r.squared),2)
```
pH predicts approx `r rsquared.ph*100` percent of the variability in yield.  

```{r, echo = FALSE}
plot(pred.ph~testGM$yield, ylab = "Predicted yield from pH model", xlab = "Actual yield from test dataset") 
abline(a = 0, b = 1)
```

## Variable selection: depth

```{r, echo = FALSE}
#hist(grassSoil$total.depth)
#summary(grassSoil$total.depth)
#quantile(grassSoil$total.depth, probs = seq(0,1,0.2))
breaks <- c(28,95,148, 170, 190, 204)
# specify interval/bin labels
tags <- c("28-95","95-148", "148-170","170-190", "180-204")
# bucketing values into bins
group_tags <- cut(grassSoil$total.depth, 
                  breaks=breaks, 
                  include.lowest=TRUE, 
                  right=FALSE, 
                  labels=tags)
#summary(group_tags)
grassSoil$depthCat <- group_tags
sum_depth <- grassSoil %>%
  group_by(depthCat) %>%
  summarize(mnYld = mean(yield, na.rm = TRUE),
            stdev = sd(yield, na.rm = TRUE),
            count = n(),
            se = stdev/sqrt(count))
#sum_depth
ggplot(sum_depth, aes(x = depthCat, y = mnYld, label = count))+
  geom_point(size = 3)+
  geom_text(vjust = -2, (aes(label = paste("n =", count))))+
  geom_errorbar(aes(ymin = mnYld - se, ymax = mnYld + se, width = 0.2))+
  ylab("Mean Yield (tons/acre) +/- se")+
  xlab("Soil depth (cm)")+
  ylim(1.5,3.5)+
  ggtitle("Grass yield by soil depth (binned)")
```

## Depth

```{r, echo = FALSE}
depth.mod <- lm(yield~total.depth, data = trainGM)
pred.depth <- predict(depth.mod, newdata = testGM)
RMSE.depth <- round(sqrt(mean((pred.depth-testGM$yield)^2)),2)
rsquared.depth <- round((summary(depth.mod)$r.squared),2)
```
Soil depth predicts approx `r rsquared.depth*100` percent of the variability in yield. 

```{r, echo = FALSE}
plot(pred.depth~testGM$yield, ylab = "Predicted yield from soil depth model", xlab = "Actual yield from test dataset") 
abline(a = 0, b = 1)
```

## Variable selection: silt

```{r, echo = FALSE}
#hist(grassSoil$silt)
#quantile(grassSoil$silt, probs = seq(0,1,0.2))
breaks <- c(0.97,20,40, 60, 79.17)
# specify interval/bin labels
tags <- c("0.97-20","20-40", "40-60","60-73.17")
# bucketing values into bins
group_tags <- cut(grassSoil$silt, 
                  breaks=breaks, 
                  include.lowest=TRUE, 
                  right=FALSE, 
                  labels=tags)
#summary(group_tags)
grassSoil$siltCat <- group_tags
sum_silt <- grassSoil %>%
  group_by(siltCat) %>%
  summarize(mnYld = mean(yield, na.rm = TRUE),
            stdev = sd(yield, na.rm = TRUE),
            count = n(),
            se = stdev/sqrt(count))
#sum_depth
ggplot(sum_silt, aes(x = siltCat, y = mnYld, label = count))+
  geom_point(size = 3)+
  geom_text(vjust = -2, (aes(label = paste("n =", count))))+
  ylim(1.5,4.3)+
  geom_errorbar(aes(ymin = mnYld - se, ymax = mnYld + se, width = 0.2))+
  ylab("Mean Yield (tons/acre) +/- se")+
  xlab("Silt percent")+
  ggtitle("Grass yield by silt percent (binned)")

```

## Silt 

```{r, echo=FALSE}
silt.mod <- lm(yield~silt, data = trainGM)
pred.silt <- predict(silt.mod, newdata = testGM)
RMSE.silt <- round(sqrt(mean((pred.silt-testGM$yield)^2)),2)
rsquared.silt <- round((summary(silt.mod)$r.squared),2)
```
Percent silt predicts approx `r rsquared.silt*100` percent of the variability in yield. 

```{r, echo = FALSE}
plot(pred.silt~testGM$yield, ylab = "Predicted yield from silt model", xlab = "Actual yield from test dataset") 
abline(a = 0, b = 1)
```

## Variable selection: sand

```{r, echo = FALSE}
#hist(grassSoil$sand)
#summary(grassSoil$sand)
breaks <- c(6,20,40, 60, 80,96.87)
# specify interval/bin labels
tags <- c("6-20","20-40", "40-60","60-80", "80-96.87")
# bucketing values into bins
group_tags <- cut(grassSoil$sand, 
                  breaks=breaks, 
                  include.lowest=TRUE, 
                  right=FALSE, 
                  labels=tags)
#summary(group_tags)
grassSoil$sandCat <- group_tags
sum_sand <- grassSoil %>%
  group_by(sandCat) %>%
  summarize(mnYld = mean(yield, na.rm = TRUE),
            stdev = sd(yield, na.rm = TRUE),
            count = n(),
            se = stdev/sqrt(count))
#sum_depth
ggplot(sum_sand, aes(x = sandCat, y = mnYld, label = count))+
  geom_point(size = 3)+
  geom_text(vjust = -2, (aes(label = paste("n =", count))))+
  geom_errorbar(aes(ymin = mnYld - se, ymax = mnYld + se, width = 0.2))+
  ylim(1.5,4.2)+
  ylab("Mean Yield (tons/acre) +/- se")+
  xlab("Sand percent")+
  ggtitle("Grass yield by sand percent (binned)")

```

## Sand

```{r, echo = FALSE}
sand.mod <- lm(yield~sand, data = trainGM)
pred.sand <- predict(sand.mod, newdata = testGM)
RMSE.sand <- round(sqrt(mean((pred.sand-testGM$yield)^2)),2)
rsquared.sand <- round((summary(sand.mod)$r.squared),2)
```
Percent sand predicts approx `r rsquared.sand*100` percent of the variability in yield. 

```{r, echo = FALSE}
plot(pred.sand~testGM$yield, ylab = "Predicted yield from sand model", xlab = "Actual yield from test dataset") 
abline(a = 0, b = 1)
```

## Variable selection: Grass type

```{r, echo = FALSE}
sum_crop <- grassSoil %>%
  group_by(cropname) %>%
  summarize(mnYld = mean(yield, na.rm = TRUE),
            stdev = sd(yield, na.rm = TRUE),
            count = n(),
            se = stdev/sqrt(count))

ggplot(sum_crop, aes(x = cropname, y = mnYld, label = count))+
  geom_point(size = 3)+
  geom_text(vjust = -2, (aes(label = paste("n =", count))))+
  geom_errorbar(aes(ymin = mnYld - se, ymax = mnYld + se, width = 0.2))+
  ylim(1.5,4.2)+
  ylab("Mean Yield (tons/acre) +/- se")+
  xlab(" ")+
  ggtitle("Grass yield by grass")

```

## Grass type

```{r,echo=FALSE}
grass.mod <- lm(yield~cropname, data = trainGM)
pred.grass <- predict(grass.mod, newdata = testGM)
RMSE.grass <- round(sqrt(mean((pred.grass-testGM$yield)^2)),2)
rsquared.grass <- round((summary(grass.mod)$r.squared),2)
```
Grass type predicts approx `r rsquared.grass*100` percent of the variability in yield.  

```{r, echo = FALSE}
plot(pred.grass~testGM$yield, ylab = "Predicted yield from grass type model", xlab = "Actual yield from test dataset") 
abline(a = 0, b = 1)
```

## Comparing variables 

```{r, echo=FALSE}
variables <- data.frame(var = c("Null", "Grass mix", "AWC", "CEC", "Depth", "Ksat", "OM", "pH", "Sand", "Silt", "Slope"),
                        Rsquared = c("Null", rsquared.grass, rsquared.awc, rsquared.cec, rsquared.depth, rsquared.ksat, rsquared.om, rsquared.ph, rsquared.sand, rsquared.silt, rsquared.slope),
                        RMSE = c(RMSE.baseline, RMSE.grass, RMSE.awc, RMSE.cec, RMSE.depth, RMSE.ksat, RMSE.om, RMSE.ph, RMSE.sand, RMSE.silt, RMSE.slope))
variables <- variables %>% 
  arrange(desc(RMSE)) 

variables <- variables %>%
  gt %>%
   tab_header(
    title = "Variables explored",
    subtitle = "single variable model predictions") %>%
  cols_label(var = "Variable",
             Rsquared = html("R<sup>2</sup>"))
  
variables
  
```

## New dataset

```{r}
grassSoil <- grassSoil %>%
  dplyr::select(c(yield, cropname, sand, silt, cec, awc, ksat, slope.r, ph, om, total.depth)) %>%
  rename(slope = slope.r)
summary(grassSoil)
```

## Check for correlations 

```{r, echo = FALSE}
ggpairs(data = grassSoil[,c(3:11)])
```

## Correlations 

* Sand and silt
* Sand/silt and ksat
* Sand/silt and cec
* Sand/silt and awc
* Available water capacity and cation exchange capacity

## Model selection processes

Create new training and testing datasets

```{r, cache=TRUE}
set.seed(0731)
inTrain <- createDataPartition(grassSoil$yield, p = 0.75, list = FALSE)
trainGM <- grassSoil[inTrain,]
testGM <- grassSoil[-inTrain,]
```

## Create full model for comparisons

```{r}
train.control <- trainControl(method = "cv", number = 10) #cross validate 10 times
full.model <- train(yield~., data = trainGM, method = "lm", trControl = train.control)
full.pred <- predict(full.model, testGM)
RMSE.full <- round(sqrt(mean((full.pred-testGM$yield)^2)),3)
adj.rsquared.full <- round((summary(full.model)$adj.r.squared),3)
rsquared.full <- round((summary(full.model)$r.squared),3)
models <- data.frame(Model = "Full Model", RMSE = RMSE.full, Rsquared = rsquared.full, adj.Rsquared = adj.rsquared.full)
models
```


## Check the variance inflation factor (VIF)

VIFs exceeding 4 warrant further investigation, while VIFs exceeding 10 are signs of serious multicollinearity requiring correction


```{r}
as.matrix(car::vif(full.model$finalModel))
```


## Remove either sand or silt

This is the most highly correlated.

```{r, echo=FALSE}
#create dataset without sand
noSand <- grassSoil %>% dplyr::select(-sand)
#create training and test sets
set.seed(0731)
inTrain <- createDataPartition(noSand$yield, p = 0.75, list = FALSE)
train.noSand <- noSand[inTrain,]
test.noSand <- noSand[-inTrain,]
# create model
noSand.mod <- train(yield~., data = train.noSand, method = "lm", trControl = train.control)
noSand.pred <- predict(noSand.mod, test.noSand)
RMSE.noSand <- round(sqrt(mean((noSand.pred-test.noSand$yield)^2)),3)
adj.rsquared.noSand <- round((summary(noSand.mod)$adj.r.squared),3)
rsquared.noSand <- round((summary(noSand.mod)$r.squared),3)

# create dataset without silt
noSilt <- grassSoil %>% dplyr::select(-silt)
#create training and test sets
set.seed(0731)
inTrain <- createDataPartition(noSilt$yield, p = 0.75, list = FALSE)
train.noSilt <- noSilt[inTrain,]
test.noSilt <- noSilt[-inTrain,]
# create model
noSilt.mod <- train(yield~., data = train.noSilt, method = "lm", trControl = train.control)
noSilt.pred <- predict(noSilt.mod, test.noSilt)
RMSE.noSilt <- round(sqrt(mean((noSilt.pred-test.noSilt$yield)^2)),3)
adj.rsquared.noSilt <- round((summary(noSilt.mod)$adj.r.squared),3)
rsquared.noSilt <- round((summary(noSilt.mod)$r.squared),3)

models <- add_row(models, Model = c("No Sand", "No Silt"), RMSE = c(RMSE.noSand, RMSE.noSilt), Rsquared = c(rsquared.noSand, rsquared.noSilt), 
                  adj.Rsquared = c(adj.rsquared.noSand,adj.rsquared.noSilt))
models_display <- models %>%
  gt()%>%
  tab_header(
    title = "Model comparisons") %>%
  cols_label(adj.Rsquared = html("Adj. R<sup>2</sup>"),
             Rsquared = html("R<sup>2</sup>"))
models_display
  
```

The data suggest the model is better with sand as a predictor, rather than silt. We continue with variable selection on a dataset without silt.

## VIF when Silt is removed

```{r}
as.data.frame(car::vif(noSilt.mod$finalModel))
```


## StepAIC variable selection

```{r, message= FALSE}
aic.model <- train(yield ~., data = train.noSand,
                    method = "lmStepAIC", 
                    trControl = train.control
                    )
```

## StepAIC variable selection

```{r}
aic.model$finalModel
summary(aic.model$finalModel) 
```

## Step AIC says take out CEC
Test the datasets with and without CEC

```{r}
# create dataset without silt
noCEC <- noSilt %>% dplyr::select(-cec)
#create training and test sets
set.seed(0731)
inTrain <- createDataPartition(noCEC$yield, p = 0.75, list = FALSE)
train.noCEC <- noCEC[inTrain,]
test.noCEC <- noCEC[-inTrain,]
# create model
noCEC.mod <- train(yield~., data = train.noCEC, method = "lm", trControl = train.control)
noCEC.pred <- predict(noCEC.mod, test.noCEC)
RMSE.noCEC <- round(sqrt(mean((noCEC.pred-test.noCEC$yield)^2)),3)
adj.rsquared.noCEC <- round((summary(noCEC.mod)$adj.r.squared),3)
rsquared.noCEC <- round((summary(noCEC.mod)$r.squared),3)

# create dataset without sand
noCEC2 <- noSand %>% dplyr::select(-cec)
#create training and test sets
set.seed(0731)
inTrain <- createDataPartition(noCEC2$yield, p = 0.75, list = FALSE)
train.noCEC2 <- noCEC2[inTrain,]
test.noCEC2 <- noCEC2[-inTrain,]
# create model
noCEC.mod2 <- train(yield~., data = train.noCEC2, method = "lm", trControl = train.control)
noCEC.pred2 <- predict(noCEC.mod2, test.noCEC2)
RMSE.noCEC2 <- round(sqrt(mean((noCEC.pred2-test.noCEC2$yield)^2)),3)
adj.rsquared.noCEC2 <- round((summary(noCEC.mod2)$adj.r.squared),3)
rsquared.noCEC2 <- round((summary(noCEC.mod2)$r.squared),3)

models <- add_row(models, Model = c("No CEC/No Silt", "No CEC/No sand"), RMSE = c(RMSE.noCEC, RMSE.noCEC2), 
                  Rsquared = c(rsquared.noCEC, rsquared.noCEC2), adj.Rsquared = c(adj.rsquared.noCEC, adj.rsquared.noCEC2))
models <- models %>%
  arrange(RMSE)
model_display <- models %>%
  gt()%>%
  tab_header(
    title = "Model comparisons") %>%
  cols_label(adj.Rsquared = html("Adj. R<sup>2</sup>"),
             Rsquared = html("R<sup>2</sup>"))

model_display
```


## Check the VIF again

```{r}
as.data.frame(car::vif(noCEC.mod2$finalModel))
```
## Silt and AWC still create a problem

Remove AWC

```{r, echo=FALSE}
# create dataset without awc or silt
noAWC <- noCEC2 %>% dplyr::select(-awc)
#create training and test sets
set.seed(0731)
inTrain <- createDataPartition(noAWC$yield, p = 0.75, list = FALSE)
train.noAWC <- noAWC[inTrain,]
test.noAWC <- noAWC[-inTrain,]
# create model
noAWC.mod <- train(yield~., data = train.noAWC, method = "lm", trControl = train.control)
noAWC.pred <- predict(noAWC.mod, test.noAWC)
RMSE.noAWC <- round(sqrt(mean((noAWC.pred-test.noAWC$yield)^2)),3)
adj.rsquared.noAWC <- round((summary(noAWC.mod)$adj.r.squared),3)
rsquared.noAWC <- round((summary(noAWC.mod)$r.squared),3)

models <- add_row(models, Model = c("No sand/CEC/AWC"), RMSE = c(RMSE.noAWC), Rsquared = c(rsquared.noAWC), 
                  adj.Rsquared = c(adj.rsquared.noAWC))
models <- models %>%
  arrange(RMSE)

model_display <- models %>%
  gt()%>%
  tab_header(
    title = "Model comparisons") %>%
  cols_label(adj.Rsquared = html("Adj. R<sup>2</sup>"),
             Rsquared = html("R<sup>2</sup>"))

model_display
```

## Remove AWC

Much better.

```{r}
as.data.frame(vif(noAWC.mod$finalModel))
```

## Test two way interactions
silt:crop

```{r, echo = FALSE}
ggplot(grassSoil, aes(y = yield, x = silt, color = cropname))+
  geom_jitter()
```

## Silt:crop

* No significant interaction
```{r}
silt.crop.mod <- lm(yield~silt*cropname, data = grassSoil)
anova(silt.crop.mod)
```

## Two way interactions
Silt:ksat


```{r, echo = FALSE}
ggplot(grassSoil, aes(y = yield, x = silt, color = ksat))+
  geom_jitter()
```

## Silt:ksat

* Significant interaction
```{r}
silt.ksat.mod <- lm(yield~silt*ksat, data = grassSoil)
anova(silt.ksat.mod)
```

## Two way interactions
Silt:slope


```{r, echo = FALSE}
ggplot(grassSoil, aes(y = yield, x = silt, color = slope))+
  geom_jitter()
```

## Silt:slope

* Significant interaction. 
```{r}
silt.slope.mod <- lm(yield~silt*slope, data = grassSoil)
anova(silt.slope.mod)
```

## Two way interactions
Silt:ph


```{r, echo = FALSE}
ggplot(grassSoil, aes(y = yield, x = silt, color = ph))+
  geom_jitter()
```

## Silt:ph

* Significant interaction. 
```{r}
silt.ph.mod <- lm(yield~silt*ph, data = grassSoil)
anova(silt.ph.mod)
```
## Two way interactions
Silt:om


```{r, echo = FALSE}
ggplot(grassSoil, aes(y = yield, x = silt, color = om))+
  geom_jitter()
```

## Silt:om

* No significant interaction. 
```{r}
silt.om.mod <- lm(yield~silt*om, data = grassSoil)
anova(silt.om.mod)
```
## Two way interactions
Silt:depth


```{r, echo = FALSE}
ggplot(grassSoil, aes(y = yield, x = silt, color = total.depth))+
  geom_jitter()
```

## Silt:depth

* Significant interaction. 
```{r}
silt.depth.mod <- lm(yield~silt*total.depth, data = grassSoil)
anova(silt.depth.mod)
```

## Two way interactions: Silt

Sand interacts with: ksat, slope, pH, OM


## Two way interactions
crop:ksat
 
```{r, echo = FALSE}
ggplot(grassSoil, aes(y = yield, x = cropname, color = ksat))+
  geom_jitter()
```

## Crop:ksat

* No significant interaction 
```{r}
crop.ksat.mod <- lm(yield~cropname*ksat, data = grassSoil)
anova(crop.ksat.mod)
```

## Two way interactions
crop:slope
 
```{r, echo = FALSE}
ggplot(grassSoil, aes(y = yield, x = cropname, color = slope))+
  geom_jitter()
```

## Crop:slope

* No significant interaction 
```{r}
crop.slope.mod <- lm(yield~cropname*slope, data = grassSoil)
anova(crop.slope.mod)
```

## Two way interactions
crop:ph
 

```{r, echo = FALSE}
ggplot(grassSoil, aes(y = yield, x = cropname, color = ph))+
  geom_jitter()
```

## Crop:ph

* No significant interaction 
```{r}
crop.ph.mod <- lm(yield~cropname*ph, data = grassSoil)
anova(crop.ph.mod)
```

## Two way interactions
crop:om
 
 
```{r, echo = FALSE}
ggplot(grassSoil, aes(y = yield, x = cropname, color = om))+
  geom_jitter()
```

## Crop:om

* No significant interaction
```{r}
crop.om.mod <- lm(yield~cropname*om, data = grassSoil)
anova(crop.om.mod)
```

## Two way interactions
crop:depth
 
 
```{r, echo = FALSE}
ggplot(grassSoil, aes(y = yield, x = cropname, color = total.depth))+
  geom_jitter()
```

## Crop:depth

* No significant interaction
```{r}
crop.depth.mod <- lm(yield~cropname*total.depth, data = grassSoil)
anova(crop.depth.mod)
```

## Two way interactions: crop

Crop doesn't interact with any variable!


## Two way interactions

ksat: slope


```{r, echo = FALSE}
ggplot(grassSoil, aes(y = yield, x = slope, color = ksat))+
  geom_jitter()
```

## Ksat:slope

* Significant interaction
```{r}
ksat.slope.mod <- lm(yield~ksat*slope, data = grassSoil)
anova(ksat.slope.mod)
```

## Two way interactions

ksat: ph


```{r, echo = FALSE}
ggplot(grassSoil, aes(y = yield, x = ph, color = ksat))+
  geom_jitter()
```

## Ksat:pH

* Significant interaction
```{r}
ksat.ph.mod <- lm(yield~ksat*ph, data = grassSoil)
anova(ksat.ph.mod)
```

## Two way interactions

ksat: om


```{r, echo = FALSE}
ggplot(grassSoil, aes(y = yield, x = total.depth, color = om))+
  geom_jitter()
```

## Ksat:om

* Significant interaction
```{r}
ksat.om.mod <- lm(yield~ksat*om, data = grassSoil)
anova(ksat.om.mod)
```

## Two way interactions

ksat: depth


```{r, echo = FALSE}
ggplot(grassSoil, aes(y = yield, x = total.depth, color = ksat))+
  geom_jitter()
```

## Ksat:depth

* Significant interaction
```{r}
ksat.depth.mod <- lm(yield~ksat*total.depth, data = grassSoil)
anova(ksat.depth.mod)
```

## Two way interactions: saturated hydraulic conductivity

Ksat interacts with: slope, pH, OM, depth.

## Two way interactions

slope: depth


```{r, echo = FALSE}
ggplot(grassSoil, aes(y = yield, x = slope, color = total.depth))+
  geom_jitter()
```

## Slope:depth

* No Significant interaction
```{r}
slope.depth.mod <- lm(yield~slope*total.depth, data = grassSoil)
anova(slope.depth.mod)
```

## Two way interactions

slope: ph


```{r, echo = FALSE}
ggplot(grassSoil, aes(y = yield, x = slope, color = ph))+
  geom_jitter()
```

## slope:pH

* Significant interaction
```{r}
slope.ph.mod <- lm(yield~slope*ph, data = grassSoil)
anova(slope.ph.mod)
```

## Two way interactions

slope: om


```{r, echo = FALSE}
ggplot(grassSoil, aes(y = yield, x = slope, color = om))+
  geom_jitter()
```

## slope:om

* Significant interaction
```{r}
slope.om.mod <- lm(yield~slope*om, data = grassSoil)
anova(slope.om.mod)
```

## Two way interactions: slope

Slope interacts with: pH and OM.

## Two way interaction

ph:depth


```{r, echo = FALSE}
ggplot(grassSoil, aes(y = yield, x = ph, color = total.depth))+
  geom_jitter()
```

## pH:depth

* Significant interaction
```{r}
ph.depth.mod <- lm(yield~ph*total.depth, data = grassSoil)
anova(ph.depth.mod)
```

## Two way interactions

pH:om


```{r, echo = FALSE}
ggplot(grassSoil, aes(y = yield, x = ph, color = om))+
  geom_jitter()
```

## pH:OM

* No sgnificant interaction
```{r}
om.ph.mod <- lm(yield~om*ph, data = grassSoil)
anova(om.ph.mod)
```

## Two way interactions: pH

pH interacts with depth

## Two way interactions

om:depth


```{r, echo = FALSE}
ggplot(grassSoil, aes(y = yield, x = total.depth, color = om))+
  geom_jitter()
```

## OM:depth

* No significant interaction
```{r}
om.depth.mod <- lm(yield~om*total.depth, data = grassSoil)
anova(om.depth.mod)
```

## Two way interactions: Organic matter

No more interactions

## Create training and test sets
Data with crop, silt, ksat, slope, pH, OM, depth (no sand, cec, or awc)

```{r}
finalDat <- noAWC
set.seed(0731)
inTrain <- createDataPartition(finalDat$yield, p = 0.75, list = FALSE)
train.final <- finalDat[inTrain,]
test.final <- finalDat[-inTrain,]
```


## Create model with significant interactions

Drop non-significant interaction terms until they are all significant, or until removing them decreases model performance.
```{r, echo=FALSE}
mod.1 <- train(yield ~ . ^2,
               data = train.final, method = "lm", trControl = train.control)
pred.mod1 <- predict(mod.1, test.final)
RMSE.mod1 <- round(sqrt(mean((pred.mod1-test.final$yield)^2)),3)
adj.rsquared.mod1 <- round((summary(mod.1)$adj.r.squared),3)
rsquared.mod1 <- round((summary(mod.1)$r.squared),3)
mod.1length <- length(coef(mod.1$finalModel))


# remove crop interactions
mod.1a <- train(yield ~ . ^2 
                - cropname:silt - cropname:ksat - cropname:slope - cropname:ph - cropname:om - 
                  cropname:total.depth ,
               data = train.final, method = "lm", trControl = train.control)
pred.mod1a <- predict(mod.1a, test.final)
RMSE.mod1a <- round(sqrt(mean((pred.mod1a-test.final$yield)^2)),3)
adj.rsquared.mod1a <- round((summary(mod.1a)$adj.r.squared),3)
rsquared.mod1a <- round((summary(mod.1a)$r.squared),3)
mod1alength <- length(coef(mod.1a$finalModel))


mod.2 <- train(yield ~ . + silt*ksat + silt*slope + silt*ph + silt*total.depth + 
                 ksat*slope + ksat*ph + ksat*total.depth + ksat*om +
                 slope*ph + slope*om +
                 ph*total.depth,
               data = train.final, method = "lm", trControl = train.control)
pred.mod2 <- predict(mod.2, test.final)
RMSE.mod2 <- round(sqrt(mean((pred.mod2-test.final$yield)^2)),3)
adj.rsquared.mod2 <- round((summary(mod.2)$adj.r.squared),3)
rsquared.mod2 <- round((summary(mod.2)$r.squared),3)
mod2length <- length(coef(mod.2$finalModel))
#summary(mod.2)

#remove ksat:total depth
mod.2a <- train(yield ~ . + silt*ksat + silt*slope + silt*ph + silt*total.depth + 
                 ksat*slope + ksat*ph + ksat*om +
                 slope*ph + slope*om +
                 ph*total.depth,
               data = train.final, method = "lm", trControl = train.control)
pred.mod2a <- predict(mod.2a, test.final)
RMSE.mod2a <- round(sqrt(mean((pred.mod2a-test.final$yield)^2)),3)
adj.rsquared.mod2a <- round((summary(mod.2a)$adj.r.squared),3)
rsquared.mod2a <- round((summary(mod.2a)$r.squared),3)
mod2alength <- length(coef(mod.2a$finalModel))
#summary(mod.2a)
noAWClength <- length(coef(noAWC.mod$finalModel))
 
InterXmodels <- data.frame(Model = c("Full InterX mod", "No crop interX", "InterX from two way tests", "Drop ksat:depth", "No 
                                     interactions", "Baseline"),
                  RMSE = c(RMSE.mod1, RMSE.mod1a, RMSE.mod2, RMSE.mod2a, RMSE.noAWC, RMSE.baseline), 
                  adj.rsquared = c(adj.rsquared.mod1, adj.rsquared.mod1a, adj.rsquared.mod2, adj.rsquared.mod2a,
                                   adj.rsquared.noAWC, "NA"),
                  Rsquared = c(rsquared.mod1, rsquared.mod1a, rsquared.mod2, rsquared.mod2a, rsquared.noAWC, "NA"),
                  NoCoef = c(mod.1length, mod1alength, mod2length, mod2alength, noAWClength, "NA"))
model.compare <- InterXmodels %>%
  gt()%>%
  tab_header(
    title = "Model comparisons") %>%
  cols_label(adj.rsquared = html("Adj. R<sup>2</sup>"),
             Rsquared = html("R<sup>2</sup>"), 
             NoCoef = "No. coefs")
model.compare  
```

## Predicted versus test values

```{r, echo = FALSE}
plot(test.final$yield, pred.mod1a, xlab = "Test data set yields", ylab = "Predicted yields from final model")
abline(a = 0, b = 1)
```
